Reducing retrieval error in MAIAC AOD with the `Earth_obs_cleaning` machine-learning workflow: Results from CONUS

### 2023-08-26 writing/CONUS_AOD.qmd
Minor text updates and add captions.

We demonstrate a reproducible workflow to reduce retrieval error from satellite-based datasets with an application to MODIS Aerosol Optical Depth over the conterminous United States 2000-2021. It facilitates the often laborious task of obtaining and comparing ground-based and remotely sensed quantities for product validation and evaluating retrieval biases. The ML data-driven approach demonstrates impressive improvements in reducing retrieval error with real-world impacts for air quality modeling.

@@

The study region was the conterminous United States (CONUS) and all match-ups of MODIS data were based on the 1-km grid cells in the Global sinusoidal Coordinate Reference System. The study period included all retrievals from `r min(Wf$dates)` through `r max(Wf$dates)`. The `Earth_obs_cleaning` workflow automatically downloaded all intersecting tiles of MCD19A2 Collection 6[@lyapustinalexei2018] based on a NASA-supplied MODIS tile boundary file. While the MCD19A2 product includes data from both Aqua and Terra MODIS instruments, we focus our analyses on the longer running Terra and report results for Aqua in Supplementary Data.

@@

Our primary metric to evaluate the difference between MAIAC and AERONET before and after applying our correction was the root mean square error (RMSE), which is at the original scale of the AOD and can be compared in magnitude with the overall variability based on the standard deviation (SD) of the AERONET AOD. We also report the SD for raw and corrected MAIAC AOD to examine how our empirical adjustments impact variability. We calculate bias (the mean difference of satellite AOD versus AERONET observations) for the raw and corrected products. We also report the Pearson correlation of the raw and corrected products versus the collocated AERONET measures. Finally, we report on the distribution of differences between the acquisition time for each overpass-tile and the nearest AERONET observation in time. We plotted the agreement of raw and corrected AOD versus AERONET AOD in scatterplots for the interval [0, 1] given that almost all AERONET observations in the cross-validation set fell within that range. To generate empirical error envelopes with performance for both low and high AOD conditions, we checked the combination of additive and multiplicative terms closest to containing \~2/3 of observations for AERONET AOD below and above 0.6.

While the model was trained using all of the collocated observations generated from within the study domain defined above, we present a number of stratified tables with the same metrics listed above to examine performance of the correction procedure in various data subsets. Stratification variables included time/space (by year, months of the year, and NOAA climate regions), atmospheric conditions (low versus high AOD), closeness of the temporal matchup (above versus below median time difference), and two additional factors related to MAIAC retrieval (number of overpasses per day and MAIAC quality flag).

@@

### Performance for higher versus lower AOD (AERONET \> 0.6)

```{r}
by.highaod = dt[, eval(performance.j), by = .(aod_high = y.ground > 0.6)]
flextable(pretty.table.numbers(by.highaod))
```

### Performance by `qa_best`

```{r}
by.qabest = dt[, eval(performance.j), by = qa_best]
flextable(pretty.table.numbers(by.qabest))
```

@@

```{r agreement_plot}
#| label: fig-agreement-plot
#| fig-cap: !expr caption
caption = with(list(d = agreement.plot.data(dt)), sprintf(
    "Not shown: %s %s and %s %s",
    d[!in.region & comparison == "y.sat", .N],
    "points outside left panel",
    d[!in.region & comparison == "y.ground.pred", .N],
    "points outside right panel"))
IG(tar_read(agreement.plot.path))
```

@@

## Comparison with AQS

We examined `r l = tar_read(satellite.vs.aqs); scales::comma(l$nobs)` observations of AQS on `r scales::comma(l$celldays)` cell-days. The correlation of the observations with the original satellite values was `r sprintf("%.03f", l$old)`, compared to `r sprintf("%.03f", l$new)` with our corrected values.

@@

Our workflow treats all cloud-screened and quality-assured (Level 2.0) observations from AERONET instruments as reference data and does not require a minimum number of days of observation from a given site for inclusion (as might be desired in an analysis of long-term MODIS sensor drift). The restriction to this highest quality subset of AERONET means that the training set may have few or no examples of tricky retrieval conditions (e.g., cloud edges). Similarly, nearly all AERONET stations are over land and thus may not adequately represent retrieval errors over water and mixed pixels, however the MAIAC algorithm itself is applicable over land and inland water (and not the global ocean). We leave for a future analysis the applicability of a trained `Earth_obs_cleaning` model to other space-time regions (with similar or different aerosol properties). Although here we present the `Earth_obs_cleaning` trained on almost 100,000 collocated observations, we use machine-learning methods likely to perform well with substantially smaller training datasets such as would be available for more recently launched sensors or over regions with more sparse ground network coverage.

>>>>

We demonstrate a reproducible workflow to reduce retrieval error from satellite-based datasets with an application to MODIS Aerosol Optical Depth over the conterminous United States 2000-2022. It facilitates the often laborious task of obtaining and comparing ground-based and remotely sensed quantities for product validation and evaluating retrieval biases. The ML data-driven approach demonstrates impressive improvements in reducing retrieval error with real-world impacts for air quality modeling.

@@

The study region was the conterminous United States (CONUS) and all match-ups of MODIS data were based on the 1-km grid cells in the Global sinusoidal Coordinate Reference System. The study period included all retrievals from `r min(Wf$dates)` through `r max(Wf$dates)`. The `Earth_obs_cleaning` workflow automatically downloaded all intersecting tiles of MCD19A2 Collection 6.1[@lyapustinalexei2018] based on a NASA-supplied MODIS tile boundary file. While the MCD19A2 product includes data from both Aqua and Terra MODIS instruments, we focus our analyses on the longer running Terra.

@@

Our primary metric to evaluate the difference between MAIAC and AERONET before and after applying our correction was the root mean square error (RMSE), which is at the original scale of the AOD and can be compared in magnitude with the overall variability based on the standard deviation (SD) of the AERONET AOD. We also report the SD for raw and corrected MAIAC AOD to examine how our empirical adjustments impact variability. We calculate bias (the mean difference of satellite AOD versus AERONET observations) for the raw and corrected products. We also report the Pearson correlation of the raw and corrected products versus the collocated AERONET measures. We plotted the agreement of raw and corrected AOD versus AERONET AOD in scatterplots for the interval [0, 1] given that almost all AERONET observations in the cross-validation set fell within that range. To generate empirical error envelopes with performance for both low and high AOD conditions, we optimized the combination of additive and multiplicative terms closest to containing \~2/3 of observations for AERONET AOD below and above 0.6, a threshold previously used when evaluating MAIAC AOD.

While the model was trained using all of the collocated observations generated from within the study domain defined above, we present a number of stratified tables with the same metrics listed above to examine performance of the correction procedure in various data subsets. Stratification variables included time/space (NOAA climate regions, months of the year, and by year), atmospheric conditions (low versus high AOD), and two additional factors related to MAIAC retrieval (MAIAC quality flag and number of overpasses per day).

@@

```{r}
#| label: tbl-cv-high-aod
#| tbl-cap: Performance for higher versus lower AOD (AERONET \> 0.6).
by.highaod = dt[, eval(performance.j), by = .(aod_high = y.ground > 0.6)]
flextable(pretty.table.numbers(by.highaod))
```

```{r}
#| label: tbl-cv-qa-best
#| tbl-cap: Performance by quality flag `qa_best`.
by.qabest = dt[, eval(performance.j), by = qa_best]
flextable(pretty.table.numbers(by.qabest))
```

@@

```{r agreement_plot}
#| label: fig-agreement-plot
#| fig-cap: !expr caption
caption = with(list(d = agreement.plot.data(dt)), sprintf(
    "Agreement plots for satellite-based and AERONET AOD. Straight lines show empirical error envelopes containing 2/3 of retrievals. Not shown: %s %s and %s %s",
    d[!in.region & comparison == "y.sat", .N],
    "points outside left panel",
    d[!in.region & comparison == "y.ground.pred", .N],
    "points outside right panel"))
IG(tar_read(agreement.plot.path))
```

@@

## Comparison with AQS PM~2.5~

We examined `r l = tar_read(satellite.vs.aqs); scales::comma(l$nobs)` observations of AQS PM~2.5~ on `r scales::comma(l$celldays)` cell-days. The correlation of the observations with the original satellite values was `r sprintf("%.03f", l$old)`, compared to `r sprintf("%.03f", l$new)` with our corrected values.

@@

Our workflow treats all cloud-screened and quality-assured (Level 2.0) observations from AERONET instruments as reference data and does not require a minimum number of days of observation from a given site for inclusion (as might be desired in an analysis of long-term MODIS sensor drift). The restriction to this highest quality subset of AERONET means that the training set may have few or no examples of tricky retrieval conditions (e.g., cloud edges). Similarly, nearly all AERONET stations are over land and thus may not adequately represent retrieval errors over water and mixed pixels, however the MAIAC algorithm itself is applicable over land and inland water (and not the global ocean). We leave for a future analysis the applicability of a trained `Earth_obs_cleaning` model to other space-time regions (with similar or different aerosol properties). Although here we present the `Earth_obs_cleaning` trained on over 100,000 collocated observations, we use machine-learning methods likely to perform well with substantially smaller training datasets such as would be available for more recently launched sensors or over regions with more sparse ground network coverage.
